---
title: "Project Final"
author: "Parv Maru"
date: "2025-05-25"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(scales)
library(ggthemes)
library(dplyr)
library(readr)
```

```{r}
# Loading all three datasets
df_2021 <- read_csv("ipums_2021_data.csv")
df_2022 <- read_csv("ipums_2022_data.csv")
df_2023 <- read_csv("ipums_2023_data.csv")

# Combining datasets
df_combined <- bind_rows(df_2021, df_2022, df_2023)

# Selecting relevant variables
df_cleaned <- df_combined %>%
  select(
    YEAR, PERWT, SEX, AGE, EDUCD, EMPSTAT, INCTOT,
    UHRSWORK, WKSWORK1, CLASSWKR,
    MARST, RACE, HHINCOME
  )

# Converting codes to labels

df_cleaned <- df_cleaned %>%
  mutate(SEX = case_when(
    SEX == 1 ~ "Male",
    SEX == 2 ~ "Female",
    SEX == 9 ~ NA_character_,
    TRUE ~ NA_character_
  ))

df_cleaned <- df_cleaned %>%
  mutate(EDUCD = case_when(
    EDUCD == 000 ~ "N/A or no schooling",
    EDUCD == 001 ~ "N/A",
    EDUCD == 002 ~ "No schooling completed",
    EDUCD == 010 ~ "Nursery school to grade 4",
    EDUCD == 011 ~ "Nursery school, preschool",
    EDUCD == 012 ~ "Kindergarten",
    EDUCD == 013 ~ "Grade 1, 2, 3, or 4",
    EDUCD == 014 ~ "Grade 1",
    EDUCD == 015 ~ "Grade 2",
    EDUCD == 016 ~ "Grade 3",
    EDUCD == 017 ~ "Grade 4",
    EDUCD == 020 ~ "Grade 5, 6, 7, or 8",
    EDUCD == 021 ~ "Grade 5 or 6",
    EDUCD == 022 ~ "Grade 5",
    EDUCD == 023 ~ "Grade 6",
    EDUCD == 024 ~ "Grade 7 or 8",
    EDUCD == 025 ~ "Grade 7",
    EDUCD == 026 ~ "Grade 8",
    EDUCD == 030 ~ "Grade 9",
    EDUCD == 040 ~ "Grade 10",
    EDUCD == 050 ~ "Grade 11",
    EDUCD == 060 ~ "Grade 12",
    EDUCD == 061 ~ "12th grade, no diploma",
    EDUCD == 062 ~ "High school graduate or GED",
    EDUCD == 063 ~ "Regular high school diploma",
    EDUCD == 064 ~ "GED or alternative credential",
    EDUCD == 065 ~ "Some college, but less than 1 year",
    EDUCD == 070 ~ "1 year of college",
    EDUCD == 071 ~ "1+ years of college credit, no degree",
    EDUCD == 080 ~ "2 years of college",
    EDUCD == 081 ~ "Associate's degree, unspecified",
    EDUCD == 082 ~ "Associate's degree, occupational",
    EDUCD == 083 ~ "Associate's degree, academic",
    EDUCD == 090 ~ "3 years of college",
    EDUCD == 100 ~ "4 years of college",
    EDUCD == 101 ~ "Bachelor's degree",
    EDUCD == 110 ~ "5+ years of college",
    EDUCD == 111 ~ "6 years of college",
    EDUCD == 112 ~ "7 years of college",
    EDUCD == 113 ~ "8+ years of college",
    EDUCD == 114 ~ "Master's degree",
    EDUCD == 115 ~ "Professional degree",
    EDUCD == 116 ~ "Doctoral degree",
    EDUCD == 999 ~ NA_character_,
    TRUE ~ NA_character_
  ))

df_cleaned <- df_cleaned %>%
  mutate(EMPSTAT = case_when(
    EMPSTAT == 0 ~ "N/A",
    EMPSTAT == 1 ~ "Employed",
    EMPSTAT == 2 ~ "Unemployed",
    EMPSTAT == 3 ~ "Not in labor force",
    EMPSTAT == 9 ~ "Unknown/Illegible",
    TRUE ~ NA_character_
  ))

df_cleaned <- df_cleaned %>%
  mutate(CLASSWKR = case_when(
    CLASSWKR == 0 ~ "N/A",
    CLASSWKR == 1 ~ "Self-employed",
    CLASSWKR == 2 ~ "Works for wages",
    CLASSWKR == 9 ~ "Unknown",
    TRUE ~ NA_character_
  ))

df_cleaned <- df_cleaned %>%
  mutate(MARST = case_when(
    MARST == 1 ~ "Married, spouse present",
    MARST == 2 ~ "Married, spouse absent",
    MARST == 3 ~ "Separated",
    MARST == 4 ~ "Divorced",
    MARST == 5 ~ "Widowed",
    MARST == 6 ~ "Never married/single",
    MARST == 9 ~ NA_character_,
    TRUE ~ NA_character_
  ))

df_cleaned <- df_cleaned %>%
  mutate(RACE = case_when(
    RACE == 1 ~ "White",
    RACE == 2 ~ "Black/African American",
    RACE == 3 ~ "American Indian or Alaska Native",
    RACE == 4 ~ "Chinese",
    RACE == 5 ~ "Japanese",
    RACE == 6 ~ "Other Asian or Pacific Islander",
    RACE == 7 ~ "Other race, nec",
    RACE == 8 ~ "Two major races",
    RACE == 9 ~ "Three or more major races",
    TRUE ~ NA_character_
  ))

df_cleaned <- df_cleaned %>%
  mutate(HHINCOME = ifelse(HHINCOME == 9999999, NA, HHINCOME))

df_cleaned <- df_cleaned %>%
  filter(HHINCOME >= 0 | is.na(HHINCOME))
df_cleaned <- df_cleaned %>%
  drop_na(
    YEAR, PERWT, SEX, AGE, EDUCD, EMPSTAT, INCTOT,
    UHRSWORK, WKSWORK1, CLASSWKR, MARST, RACE, HHINCOME
  )
# Cleaning income fields: remove top-coded and invalid values
df_cleaned <- df_cleaned %>%
  mutate(
    HHINCOME = ifelse(HHINCOME == 9999999 | HHINCOME < 0, NA, HHINCOME),
    INCTOT   = ifelse(INCTOT == 9999999 | INCTOT < 0, NA, INCTOT)
  )

# clean UHRSWORK and WKSWORK1 to drop zero or invalid hours
df_cleaned <- df_cleaned %>%
  mutate(
    UHRSWORK = ifelse(UHRSWORK <= 0, NA, UHRSWORK),
    WKSWORK1 = ifelse(WKSWORK1 <= 0, NA, WKSWORK1)
  )

# Dropping all rows with any NA in the selected variables
df_cleaned <- df_cleaned %>%
  drop_na(
    YEAR, PERWT, SEX, AGE, EDUCD, EMPSTAT, INCTOT,
    UHRSWORK, WKSWORK1, CLASSWKR, MARST, RACE, HHINCOME)

# Saving the cleaned dataset
write_csv(df_cleaned, "ipums_final_cleaned_2021_2023.csv")
```

```{r load-data}
# Load data
data <- read.csv("ipums_final_cleaned_2021_2023.csv")
# Create derived columns
# Hourly wage
data <- data %>%
  mutate(hourly_wage = ifelse(UHRSWORK > 0 & WKSWORK1 > 0,
                              INCTOT / (UHRSWORK * WKSWORK1), NA))

# Age group
data <- data %>%
  mutate(age_group = case_when(
    AGE < 25 ~ "16–24",
    AGE < 45 ~ "25–44",
    AGE < 65 ~ "45–64",
    TRUE     ~ "65+"
  ))
```




## Research Q1 - Research Question 1: How does educational attainment affect income and employment?
```{r}
# Filter only employed individuals
employed_data <- data %>% 
  filter(EMPSTAT == "Employed", INCTOT > 0)

# Calculate mean income by education
edu_income_summary <- employed_data %>%
  group_by(EDUCD) %>%
  summarise(mean_income = mean(INCTOT, na.rm = TRUE)) %>%
  arrange(desc(mean_income)) %>%
  slice(1:5)  # Top 5 education levels by income

# Plot
ggplot(edu_income_summary, aes(x = reorder(EDUCD, mean_income), y = mean_income)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = paste0("$", round(mean_income, 0))), 
            vjust = -0.5, size = 3.5) +
  labs(title = "Top 5 Education Levels by Average Income (Employed Only)",
       x = "Education Level", y = "Average Total Income (USD)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

```




## 1. Employment Rate for Top 5 Education Levels by Income
```{r}
# Calculate employment rate and income for top 5 education levels
edu_summary <- data %>%
  filter(INCTOT > 0, !is.na(EDUCD), !is.na(EMPSTAT)) %>%
  group_by(EDUCD) %>%
  summarise(
    mean_income = mean(INCTOT, na.rm = TRUE),
    employment_rate = mean(EMPSTAT == "Employed", na.rm = TRUE),
    count = n()
  ) %>%
  arrange(desc(mean_income)) %>%
  slice(1:5)  # Take top 5 by income

# Plot
ggplot(edu_summary, aes(x = reorder(EDUCD, employment_rate), y = employment_rate)) +
  geom_line(group = 1, color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(title = "Employment Rate for Top 5 Education Levels by Income",
       x = "Education Level", y = "Employment Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

```


## Research Q2: How does gender influence employment access over time 
## 2(A). Employment Rate by Gender Over Time
```{r}
emp_gender_year <- data %>%
  mutate(employed_flag = ifelse(EMPSTAT == "Employed", 1, 0)) %>%
  group_by(YEAR, SEX) %>%
  summarise(employment_rate = mean(employed_flag, na.rm = TRUE))

ggplot(emp_gender_year, aes(x = YEAR, y = employment_rate, color = SEX)) +
  geom_point(size = 3) +
  geom_line(size = 1.5, aes(group = SEX), stat = "smooth", method = "loess", se = FALSE, span = 1) +
  scale_x_continuous(breaks = c(2021, 2022, 2023)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Employment Rate by Gender Over Time",
       x = "Year", y = "% Employed", color = "Gender") +
  theme_minimal()
```

## 2B. Average Income by Age (All Years Combined)
```{r}
income_by_age <- data %>%
  group_by(AGE) %>%
  summarise(mean_income = mean(INCTOT, na.rm = TRUE))

ggplot(income_by_age, aes(x = AGE, y = mean_income)) +
  geom_line(color = "darkblue", size = 1.2,linetype = "solid") +
  labs(title = "Average Total Income by Age",
       x = "Age", y = "Mean Total Income (USD)") +
  scale_y_continuous(labels = dollar_format()) +
  theme_minimal()
```

## 3. Household Income by Marital Status
```{r}
marital_income <- data %>%
  group_by(MARST) %>%
  summarise(mean_income = mean(HHINCOME, na.rm = TRUE))

ggplot(marital_income, aes(x = reorder(MARST, mean_income), y = mean_income, fill = MARST)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Average Household Income by Marital Status",
       x = "Marital Status", y = "Mean Household Income") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Research Q3: Do More Work Hours Always Lead to Higher Individual Income?
## 4. Income vs Weekly Hours Worked
```{r}
income_hours <- data %>%
  filter(!is.na(hourly_wage) & hourly_wage < 500) %>%
  group_by(UHRSWORK) %>%
  summarise(mean_income = mean(INCTOT, na.rm = TRUE)) %>%
  filter(UHRSWORK >= 10 & UHRSWORK <= 60)

ggplot(income_hours, aes(x = UHRSWORK, y = mean_income)) +
  geom_point(color = "purple", alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE, color = "purple", linewidth = 1.3) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "How Weekly Hours Worked Affects Total Income",
    x = "Usual Weekly Hours Worked",
    y = "Mean Total Income (USD)"
  ) +
  theme_minimal()
```

## Research Question 4: What factors are most useful for predicting income group and explaining income variation?
# Log Transformation Linear Regression with Coefficient Plot
```{r}
# Load necessary packages
library(dplyr)
library(ggplot2)


# Filter and prepare data
data <- data %>%
  filter(INCTOT > 0 & INCTOT < 300000, !is.na(SEX), !is.na(UHRSWORK), !is.na(EDUCD)) %>%
  mutate(
    SEX = as.factor(SEX),
    EDU_SIMPLE = case_when(
      EDUCD %in% c("No schooling completed", "Nursery school, preschool", "Kindergarten") ~ "No Schooling",
      EDUCD %in% c("Grade 1", "Grade 2", "Grade 3", "Grade 4") ~ "Grades 1–4",
      EDUCD %in% c("Grade 5", "Grade 6") ~ "Grades 5–6",
      EDUCD %in% c("Grade 7", "Grade 8") ~ "Grades 7–8",
      EDUCD %in% c("Grade 9", "Grade 10", "Grade 11") ~ "Grades 9–11",
      EDUCD %in% c("12th grade, no diploma", "Regular high school diploma", "GED or alternative credential") ~ "High School",
      grepl("Some college", EDUCD) ~ "Some College",
      grepl("Associate", EDUCD) ~ "Associate Degree",
      grepl("Bachelor", EDUCD) ~ "Bachelor's",
      grepl("Master|Doctoral|Professional", EDUCD) ~ "Grad/Prof Degree",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(EDU_SIMPLE))

data$EDU_SIMPLE <- factor(data$EDU_SIMPLE,
  levels = c("No Schooling", "Grades 1–4", "Grades 5–6", "Grades 7–8",
             "Grades 9–11", "High School", "Some College", "Associate Degree",
             "Bachelor's", "Grad/Prof Degree"))

# Log-transform income to reduce skew
data$log_income <- log(data$INCTOT)

# Run linear model
model <- lm(log_income ~ EDU_SIMPLE + AGE + SEX + UHRSWORK, data = data)

summary(model)


# Extract and clean coefficients table
coefs <- summary(model)$coefficients
coef_df <- data.frame(
  term = rownames(coefs),
  estimate = coefs[, "Estimate"],
  std_error = coefs[, "Std. Error"],
  p_value = coefs[, "Pr(>|t|)"]
)
rownames(coef_df) <- NULL

# Display coefficients (text output)
print(coef_df)

# Plot: Coefficient Effects (excluding intercept)
coef_plot_df <- coef_df %>%
  filter(term != "(Intercept)") %>%
  mutate(term = gsub("EDU_SIMPLE", "", term)) %>%
  arrange(estimate)

ggplot(coef_plot_df, aes(x = reorder(term, estimate), y = estimate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(estimate, 2)), vjust = -0.4, size = 3) +
  labs(title = "Effect of Variables on Log Income",
       x = "Variable", y = "Coefficient Estimate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

```





## Decision Tree
```{r}
# Load required packages
library(dplyr)
library(rpart)
library(rpart.plot)


# Clean and prepare the data
data <- data %>%
  filter(INCTOT > 0 & INCTOT < 300000) %>%
  mutate(
    IncomeGroup = cut(INCTOT,
                      breaks = c(0, 50000, 100000, 300000),
                      labels = c("Low", "Mid", "High")),
    EDU_SIMPLE = case_when(
      EDUCD %in% c("No schooling completed", "Nursery school, preschool", "Kindergarten") ~ "No Schooling",
      EDUCD %in% c("Grade 1", "Grade 2", "Grade 3", "Grade 4") ~ "Grades 1–4",
      EDUCD %in% c("Grade 5", "Grade 6") ~ "Grades 5–6",
      EDUCD %in% c("Grade 7", "Grade 8") ~ "Grades 7–8",
      EDUCD %in% c("Grade 9", "Grade 10", "Grade 11") ~ "Grades 9–11",
      EDUCD %in% c("12th grade, no diploma", "Regular high school diploma", "GED or alternative credential") ~ "High School",
      grepl("Some college", EDUCD) ~ "Some College",
      grepl("Associate", EDUCD) ~ "Associate Degree",
      grepl("Bachelor", EDUCD) ~ "Bachelor's",
      grepl("Master|Doctoral|Professional", EDUCD) ~ "Grad/Prof Degree",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(EDU_SIMPLE), !is.na(SEX), !is.na(UHRSWORK), !is.na(AGE))

# Set factor levels
data$SEX <- as.factor(data$SEX)
data$EDU_SIMPLE <- factor(data$EDU_SIMPLE,
                          levels = c("No Schooling", "Grades 1–4", "Grades 5–6", "Grades 7–8", "Grades 9–11",
                                     "High School", "Some College", "Associate Degree", "Bachelor's", "Grad/Prof Degree"))
data$IncomeGroup <- factor(data$IncomeGroup, levels = c("Low", "Mid", "High"))

# Build the decision tree
tree_model <- rpart(IncomeGroup ~ EDU_SIMPLE + AGE + SEX + UHRSWORK,
                    data = data, method = "class", cp = 0.002, maxdepth = 4)

# Plot with valid color palette for multiclass classification
rpart.plot(tree_model,
           extra = 104,
           box.palette = list("Reds", "Greys", "Greens"),  # Low = red, Mid = grey, High = green
           main = "Income Group Prediction Tree (Education, Age, Gender, Hours)")



```

#Predicting the Accuracy of the Model
```{r}

# Split into training and testing sets
set.seed(123)
sample_index <- sample(1:nrow(data), size = 0.7 * nrow(data))
train_data <- data[sample_index, ]
test_data <- data[-sample_index, ]

# Rebuild model on training set
tree_model <- rpart(IncomeGroup ~ EDU_SIMPLE + AGE + SEX + UHRSWORK,
                    data = train_data, method = "class", cp = 0.002, maxdepth = 4)

# Predict on test set
predictions <- predict(tree_model, newdata = test_data, type = "class")

# Create confusion matrix
conf_mat <- table(Predicted = predictions, Actual = test_data$IncomeGroup)

# Print confusion matrix
print(conf_mat)

# Calculate accuracy
accuracy <- sum(diag(conf_mat)) / sum(conf_mat)
cat("Accuracy:", round(accuracy * 100, 2), "%\n")

```